Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-03-05 mop/experiment.tex
updating on GitHub

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. \kevin{TODO} We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval. 
\kevin{Quantile regression for estimating the max/the envelope of the likelihood evaluations from AD, then fit MCAP/MQLE/MSLE?}

@@

\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble}

@@

\kevin{Method of Common Random Numbers, Coupling}

@@

\paragraph{Intuition:} We can justify this assumption as follows. For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded. The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are.

@@

@@

\kevin{rewrite to de-emphasize dop as in presentation, sell mop as the big theoretical idea that explains AD through PFs and unifies two previous estimators}

@@

We claim that this yields a particle filter that targets the posterior and recovers the derivative estimate of \citet{doucet2011sf, scibior2021dpf} when $\theta$ is evaluated at $\phi$. The proof can be found in Appendix \ref{app:dop}.

\begin{prop}[Correctness of Doubly Off-Policy Particle Filter Corrections]
    \label{prop:dop-correctness}

    Let the system evolve according to $\phi \in \Theta,$ that is, with the process model $\process(\cdot | \phi)$ and resampling with probabilities proportional to $f_{Y_n|X_n}(y_n^* | x_{n,j}^{P,\phi})$. Also let $\theta \in \gN(\phi)$. Then, the following corrections in the particle weights,
    \begin{equation}
    s_{n,j}=\frac{g_{n,j}^\theta}{g_{n,j}^{\phi}}=\frac{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P, \phi}; \theta)}{f_{Y_n|X_n}(y_n^*|x_{n,j}^{P,\phi}; \phi)}, \qquad r_{n,j}=\frac{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \theta)}{f_{X_n|X_{n-1}}(x_{n,j}^{P, \phi}|x_{n-1,j}^{F, \phi}; \phi)},
    \end{equation}
    yield a particle filter that is unbiased for the likelihood, even when $\theta \neq \phi$.

    In particular, when $\theta$ is evaluated at $\phi$, its gradient is the same as the estimator of \citet{doucet2011sf},
    \begin{equation}
        \nabla_\theta \hat\ell(\theta) = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{X_{0:N}, Y_{1:N}}(x_{1:N,j}^{A, F,\theta}, y_{1:N}^*).
    \end{equation}
\end{prop}

\kevin{Cut down on DOP discussion in actual manuscript, remove any references to this as an algorithm, just talk about the idea of an off-policy system/particle filter. Bottom paragraph is the punchline.}

@@

\kevin{Develop theory for mixing-time swapping? Detect when IF2 reaches a neighborhood of the MLE, by detecting whether it has reached a local sort of stationary distribution when the perturbations are lower bounded, e.g. the autocorrelation between the mean vectors to estimate the mixing time or Gelman-Rubin. In practice, just swap when IF2 converges according to the diagnostics. Theory doesn't care, you just need to in a basin of attraction that is strongly convex.}

@@

Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction. We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$. Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$. The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$, as we show in Proposition \ref{prop:dop-correctness}. As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff. We elaborate on this below.

@@

\kevin{This isn't a smoother, because we'd need future information.}

@@

    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. In this case, its gradient when $\alpha=1$ is the same as the estimator of \citet{doucet2011sf} when applied to the bootstrap filter,
    \begin{equation}
        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
    \end{equation}

@@

    that as \cite{scibior2021dpf} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter.
\end{prop}

@@

\kevin{Take derivative of loglik with regard to alpha?}
\kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias. }
\kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
\kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering. calculate bias and variance online during filtering, and update alpha accordingly. }

\kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }

\kevin{Elbow method on variance, since bias is hard to compute? }
\kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }

@@

\begin{thm}[Convergence of Gradient Methods]
\label{thm:convergence}

@@

and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
\end{thm}

@@

To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2. This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity. We write $M(t)$ for the cholera deaths in each month. As in \cite{ionides15}:

@@

\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example. This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.

@@

\begin{lem}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times. Write
    \[
    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
    \]
    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
\end{lem}

@@

\begin{lem}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
\end{lem}
\begin{proof}
    We want to show that, for any measurable bounded $h$, 
     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
    \begin{equation}\label{eq:lemma2:g}
    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
    \end{equation}
    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$. Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
    \begin{eqnarray*}        
    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
    \right]
    \\
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
    \end{eqnarray*}
    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$

@@

\begin{lem}[Particle Posteriors]
    \label{lem:posterior-proper-weights}
    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$. Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
\end{lem}

@@

    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows. Recursively applying Lemmas \ref{lem:change-measure-proper-weights} \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that 
    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
    the MOP-1 filter targets the posterior.
    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
    \[
    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
    \]
    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
\end{proof}

@@

\begin{lem}[Likelihood Proper Weighting]
    \label{lem:lik-proper-weight}
  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
\begin{equation}\label{L1}
L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
or by the after-resampling estimate,
\begin{equation}\label{L2}
L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
\end{equation}
where $L^\phi_n$ is as defined in the various algorithms.
\end{lem}

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ is properly weighted for $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{lem:change-measure-proper-weights}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).

@@

@@

The proof for Proposition \ref{prop:dop-correctness} follows from the below two lemmas. Evaluating the off-policy likelihood is possible with a weighted particle filter by using resampling weight $p(y^*_n|x_n;\phi)$ and making two corrections in the particle weight, through multiplying $w_{t,j}$ by the two likelihood ratios below,

\begin{equation}
    \label{eq:dmeas-ratio}
    s=p(y^*_n|x_n;\theta)/p(y^*_n|x_n;\phi)
\end{equation}
and
\begin{equation}
    \label{eq:rproc-ratio}
    r=p(x_n|x_{n-1};\theta)/p(x_n|x_{n-1};\phi).
\end{equation}

The following result showcases the correctness of corrections \ref{eq:dmeas-ratio} and \ref{eq:rproc-ratio}.

\begin{lem}
    For $\theta$, $\phi$ that satisfy Assumption \ref{assump:smooth-nbhd}, the following correction in the particle filter weights,
    \begin{equation}
        w_{t,j} := \tilde{w}_{t,j} \cdot p(y^*|x_{t,j}^P,\phi)\cdot r \cdot s
    \end{equation}
    successfully recovers $\lik(\theta)$.
\end{lem}

\begin{proof}
    Recall $\lik(\theta) = \E[\lik(\theta,\phi,\Omega, J)]=\int_\Omega \lik(\theta,\phi,\omega, J)\, d\mu(\omega)$ and $\lik(\theta) = \int_\Omega \lik(\theta, \omega, J) d\mu(\omega)$. We want to evaluate
    \begin{align*}
        \lik(\theta,\omega, J)
        &= \E\left[ \prod_{t=1}^T \lik_t(\theta) \middle| \omega\right]
        = \E\left[ \prod_{t=1}^T p(y_t^*|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
        &= \E\left[ \prod_{t=1}^T \frac{1}{J} \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|y_{1:t-1}^*,\theta) \middle| \omega\right] \\
        &= \E\left[ \prod_{t=1}^T\frac{1}{J}  \sum_{j=1}^J  p(y_t^*|x_{t,j}^P,\theta) p(x_{t,j}^P|x_{t-1}^F,\theta) \middle| \omega\right] 
    \end{align*}
    where $p(x_t^P|y_{1:t-1}^*,\theta)$ is the prediction distribution of particles, and the filtering distribution of particles is proportional to $p(y_t^*|x_{t,j}^P,\theta)$, the likelihood of particle $j$ at time $t$.

    Consider first the case with no resampling. Let $\lik^j(\theta)$, $\lik^j(\phi)$ be the likelihood of trajectory $j$ under $\theta$ and $\phi$ respectively. As the prediction and filtering particles are then the same, we use $x_{t,j}$ to refer to both. As such, conditional on $\omega$,
    \begin{align*}
        \lik^j(\theta) 
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \theta) p(x_{t,j}|x_{t-1,j}, \theta) \\
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \frac{p(y_t^*|x_{t,j}, \theta)}{p(y_t^*|x_{t,j}, \phi)} p(x_{t,j}|x_{t-1,j}, \theta) \frac{p(x_{t,j}|x_{t-1,j}, \theta) }{p(x_{t,j}|x_{t-1,j}, \phi) } \\
        &= \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r
    \end{align*}

    and therefore, if trajectories were proposed with equal probability with particles conditional on $\omega$,
    $$ \lik(\theta, \omega, J) = \frac{1}{J} \sum_{j=1}^J \prod_{t=1}^T p(y_t^*|x_{t,j}, \phi) \cdot s \cdot p(x_{t,j}|x_{t-1,j}, \phi)\cdot r $$
    The correction is therefore sufficient for recovering $\lik(\theta)$, with trajectories collected with the process model at $\phi$ and measurements evaluated with the measurement model at $\phi$.

    Consider now the case with resampling. Without loss of generality we can resample every timestep. Here, $x_{t,j}^F(\phi) \sim p(y^*|x_{t,j}^P, \phi)$ and $x_{t,j}^P(\phi) \sim p(x_t|x_{t-1}^F,\phi)$, and similarly for $\theta$. We use this notation to emphasize that \textbf{both} the measurement model and the previously drawn particles depend on $\phi$ or $\theta$.

>>>>

While the Monte-Carlo Adjusted Profile (MCAP) from \citet{Ionides_mcap} provides a profile likelihood confidence interval that takes into account both Monte Carlo profile error and statistical uncertainty in the likelihood, it is computationally inefficient. We provide a method for incorporating the MCAP methodology (with the poor man's likelihood profile) in our IFAD algorithm that provides a more computationally efficient profile likelihood confidence interval.

@@

@@

@@

\paragraph{Intuition:} We can justify this assumption as follows. For suitably nearby $\theta$ and $\phi$, the resampling indices for $\lik(\theta,\phi,\omega, J)$ and $\lik(\phi,\phi,\omega, J)$ are the same, eliminating any discontinuities from resampling, and that for small enough $\mathcal{N}(\phi)$ the likelihood ratios are bounded. The likelihood only changes by a factor of the likelihood ratios, which are bounded and smooth in $\theta$ if the densities used in the calculation are. For DOP, you need smooth likelihood ratios for the process model. For MOP, you need $k$ continuous derivatives from the reparametrization trick.

@@

We need the following assumption to ensure that the weighted particle filter with off-policy resampling targets the posterior:
\begin{aspt}[Bounded And Differentiable Measurement Density Ratios]
    For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{Y_n|X_n}(y_n^*|x, \theta)}{f_{Y_n|X_n}(y_n^*|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x \in \gX$ and all $\theta\in\gN(\phi)$. 
\end{aspt}

In the following assumption, the first condition is sufficient for DOP, and the second for MOP:
\begin{aspt}[Bounded And Differentiable Process Model]
    Either one of the following two conditions hold:
    \begin{enumerate}
        \item For any $\phi \in \Theta$, there exists an open neighborhood $\gN(\phi)$ containing $\phi$ such that the density ratios $\frac{f_{X_n|X_{n-1}}(x'|x, \theta)}{f_{X_n|X_{n-1}}(x'|x, \phi)}$ are bounded and have more than two continuous derivatives for all timesteps $n=1,...,N$, $x, x' \in \gX$ and all $\theta\in\gN(\phi)$, or
        \item There exists a differentiable simulator with more than two continuous derivatives for the process model. 
    \end{enumerate}
\end{aspt}

@@

@@

@@

@@

\kevin{Turn all theorems and lemmas into remarks, one big section on results we think may be true but leave to future work to prove. }

\kevin{Release arxiv version concurrently with an initial release of the package. Target April release, grant results for DMS come back then. }

\kevin{For a software paper, JOSS, signal that the software is careful. JSS is higher impact, but difficult to get a paper in. }

Observe that at Algorithm \ref{alg:mop}, at each timestep the corrected filtering importance weights $w_{n,j}^{F,\theta}$ optionally propagate to the next timestep for us to maintain the correction. We introduce a discounting parameter $\alpha$ that affects the weights, so $w_{n,j}^{P,\theta} := (w_{n-1,j}^{F,\theta})^{\alpha}$. Note that no matter what value $\alpha$ takes on, the system evolves according to $\phi$ when we perform the correction to evaluate the likelihood at $\theta$. The filter fully targets the posterior when $\theta=\phi$, or $\alpha=1$. As previously mentioned, the value of $\alpha$ therefore informs a bias-variance tradeoff. We elaborate on this below.

@@

@@

    In particular, when $\theta$ is evaluated at $\phi$, the likelihood estimate agrees with the bootstrap filter. In this case, its gradient when $\alpha=1$ is the same as the DPF estimator of \citet{scibior21} when applied to the bootstrap filter,
    \begin{equation}
        \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{0:N}}(y_{1:N}^* | x_{1:N,j}^{A, F,\theta}),
    \end{equation}

@@

    that as \cite{scibior21} show corresponds to differentiating through the fixed-seed likelihood estimate yielded by a vanilla particle filter, dropping resampling terms.

    When $\alpha=1$, the gradient estimate MOP-1 provides is consistent for the score.
\end{prop}

@@

% \kevin{Take derivative of loglik with regard to alpha?}
% \kevin{Talk about regime change the moment you move away from alpha=1, variance going from N2 to N, and bias going from 0 to something, but try to prove that decrease in variance outweights increase in bias. }
% \kevin{Create tools for choosing alpha via binary search CV from an averaged alpha=1 target}
% \kevin{suppose at each filtering step, look at gradient calculated at alpha=0, dynamically select alpha during filtering. calculate bias and variance online during filtering, and update alpha accordingly. }

% \kevin{No need to recompute so many particle filters! Just take an array of weight copies, one for each alpha. Gradients for each alpha come free the moment you compute one of them. }

% \kevin{Elbow method on variance, since bias is hard to compute? }
% \kevin{Get variance estimate, bias estimate, construct confidence intervals on bias estimate, use that to get unbiased estimate of the MSE. }

@@

\begin{conj}[Convergence of Gradient Methods]
\label{thm:convergence}

@@

and the algorithm eventually terminates after we obtain $||\nabla_\theta \ell(\theta_n)|| \leq (1+\sigma) \epsilon$.
\end{conj}

@@

To compare the performance of IFAD against that of IF2, we fit the very same model first proposed in \cite{king08} that \cite{ionides15} use to demonstrate the capabilities of IF2. This is becoming a standard benchmark, as \cite{wood16} and \cite{wycoff2024voronoi} include in their papers. This is a stochastic SIR compartmental model with transition and measurement uncertainty, where the population at time $t$, $H(t)$, is divided into the susceptible compartment $S(t)$, infected compartment $I(t)$, and three ($k=3$) recovered compartments $R_1(t), ..., R_k(t)$ denoting varying degrees of cholera immunity. We write $M(t)$ for the cholera deaths in each month. As in \cite{ionides15}:

@@

\kevin{Rewrite this para}
\paragraph{Differences Between IFAD-0 and IFAD-1:} The differences between IFAD-1 and IFAD-0 (MOP-1 and MOP-0) do not seem very pronounced in this example. This could be due to a number of reasons, such as (1) future case counts $y_{{n+1}:N}^*$ not providing a large amount of information on state identification $x_n$ given current and past measurements $y_{0:n}^*$ as mentioned in \cite{corenflos21}, and (2) the gradient normalization making the trajectories given by MOP-0 and MOP-1 more similar than they would be otherwise.

@@

\begin{conj}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Now, let $\{(Y_j^J,V_j^J),j=1,\dots,J\}$ be a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is represented, on average, $\pi^J_j J$ times. Write
    \[
    (Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big),
    \]
    where $a$ is called the ancestor function. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j),j=1,\dots,J\}$ targets $f_X$.
\end{conj}

@@

\begin{conj}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$ where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}$ targets $f_Z$.
\end{conj}
\begin{proof}
    We want to show that, for any measurable bounded $h$, 
     $$\frac{\sum_j h(\tilde Z_j^J) \, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[h(Z)] = \E\big[\E[h(Z) | X]\big].$$
    By hypothesis, for any measurable and bounded functional $g$ with domain $\gX$,
    \begin{equation}\label{eq:lemma2:g}
    \frac{\sum_j g(\tilde X_j^J)\, U_j^J}{\sum_j U_j^J} \stackrel{a.s.}{\to} \E[g(X)].
    \end{equation}
    Let $\bar{U}_j^J = \frac{J U_j^J}{\sum_j U_j^J}$. Examine the numerator and denominator of the quantity $$\frac{J^{-1}\sum_j h(\tilde Z_j^J) \, U_j^J}{J^{-1}\sum_j U_j^J} = \frac{J^{-1}\sum_j h(\tilde Z_j^J) \, \bar{U}_j^J}{J^{-1}\sum_j \bar{U}_j^J}.$$
    The denominator converges to $1$ almost surely. The numerator, on the other hand, is
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J,$$
    by the same fourth moment argument to the above lemma converges almost surely to the limit of its expectation
    \begin{eqnarray*}        
    \lim_{J\to\infty} \E\left[\frac{1}{J}\sum_j h(\tilde Z_j^J)\, \bar{U}_j^J\right] 
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  
      \E\left[ h(\tilde Z_j^J)\, \bar{U}_j^J\big|\tilde X_j^J, \bar U_j^J\right]
    \right]
    \\
    &=& \lim_{J\to\infty} \E \left[\frac{1}{J}\sum_j  \E\big[h(Z) \big| X=\tilde X_j^J \big] \, \bar U_j^J\right].
    \end{eqnarray*}
    Applying (\ref{eq:lemma2:g}) with $g(x) = \E\big[h(Z) | X=x\big]$, the average on the right hand size converges almost surely to $\E\big\{\E[h(Z)|X]\big\}=\E[h(Z)].$
    It remains to swap the limit and expectations. We can do so with the bounded convergence theorem, and therefore obtain   
    $$\frac{1}{J}\sum_j h(\tilde Z_j^J)\,\bar{U}_j^J \stackrel{a.s.}{\to} \E[h(Z)].$$

@@

\begin{conj}[Particle Posteriors]
    \label{lem:posterior-proper-weights}
    Suppose that $\{(X_j^J,u_j^J),j=1,\dots,J\}$ targets $f_X$. Also suppose that $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$. Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}$ targets $f_{X|Z}(\cdot | z^*)$.
\end{conj}

@@

    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows. Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that 
    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
    the MOP-1 filter targets the posterior.
    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
    \[
    \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
    \]
    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
\end{proof}

@@

\begin{conj}[Likelihood Proper Weighting]
    \label{lem:lik-proper-weight}
  $f_{Y_n|Y_{1:n-1}}(y_n^*|y_{1_n-1}^*;\theta)$ is consistently estimated by either the before-resampling estimate,
\begin{equation}\label{L1}
L_n^{B,\theta} =  \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
or by the after-resampling estimate,
\begin{equation}\label{L2}
L_n^{A,\theta} = L^\phi_n \frac{\sum_{j=1}^Jw^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}.
\end{equation}
where $L^\phi_n$ is as defined in the various algorithms.
\end{conj}

Here, (\ref{L1}) is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
To see  (\ref{L2}),
we write the numerator of (\ref{lem:change-measure-proper-weights}) as
\[
L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
= L^\phi_n \sum_{j=1}^J w_{n,j}^{FC,\theta} \frac{g^\phi_{n,j}}{L_n^\phi}
\]
Using Lemma \ref{lem:change-measure-proper-weights}, we resample according to probabilities $\frac{g^\phi_{n,j}}{L_n^\phi}$ to see this is properly estimated by
\[
L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},
\]
from which we obtain (\ref{L2}).

@@

\kevin{IS THIS UNBIASED?}

@@

We present the doubly off-policy differentiable particle filter for completeness. 
Here, we consider a doubly off policy particle filter, meaning that both \code{rprocess} and \code{dmeasure} are computed at $\phi$ and particles are reweighted to correspond to $\theta$.
In the event where a seed is fixed, all expressions in Algorithm \ref{alg:dop} can be considered to be conditional on $\omega \in \Omega$.
